The Gamma: Data Exploration through Iterative Prompting

### 2020-09-15 wip-design/paper.tex
Finish design, tweaks in case studies

% extend autocomplete into a full-blown interactive programming mechanism
% and show that it is viable for small programming tasks like data exploration
%
% --- remove some of the motivation (makes ppl ask bad questions)

%
% Iterative prompting - should we call it expression completion
% (is it more general result, or is it a new thing?)
% -> emphasize the why this is called differently paragraph
%
% Explain type providers a bit more (we say what it does, but not much about what
% it is - give dynamic typing in a static language)
%
% The 'then' operation was mysterious - can we add one line explanation to the operations?
% (yes we can and we should totally do that)

% \usepackage{xspace}
\usepackage{fancyvrb}
% \usepackage{inconsolata}
\usepackage{xcolor}
% \usepackage{tikz}
% \newcommand*{\priority}[1]{\begin{tikzpicture}[scale=0.12]%
%     \draw (0,0) circle (1);
%     \fill[fill opacity=1,fill=black] (0,0) -- (90:1) arc (90:90-#1*3.6:1) -- cycle;
%     \end{tikzpicture}}
% \newcommand*{\priorityc}[1]{
% \resizebox{0.8em}{0.8em}{
%   {\protect\tikz{ \protect
%     \draw[line width=1mm, black] (0,0) circle (1); \fill[fill opacity=1,fill=black] (0,0) -- (90:1) arc (90:90-#1*3.6:1) -- cycle;
%     } }}}
%
\definecolor{kvdclr}{rgb}{0.3,0.0,0.8}
\DefineVerbatimEnvironment{thegamma}{Verbatim}{fontfamily=zi4,numbers=left,xleftmargin=6mm,fontsize=\small,commandchars=\\\{\}}
% %\newcommand{\kvd}[1]{\textcolor{kvdclr}{#1}}
\newcommand{\kvd}[1]{\textbf{#1}}
\newcommand{\ikvd}[1]{{\fontfamily{zi4}\selectfont\small #1}}
% \usepackage{balance}       % to better equalize the last page
% \usepackage{graphics}      % for EPS, load graphicx instead
% \usepackage[T1]{fontenc}   % for umlauts and other diaeresis
% \usepackage{txfonts}
% \usepackage{mathptmx}
% \usepackage[pdflang={en-US},pdftex]{hyperref}
% \usepackage{color}
% \usepackage{booktabs}
% \usepackage{textcomp}

@@

\begin{figure}
\includegraphics[width=.6\columnwidth]{figures/thegamma-annot}
\caption{Obtaining teams with the greatest number of gold medals from Rio 2016
Olympics with a reproducible The Gamma script (1), contextual iterative prompting mechainsm
offering ways of sorting the data (2) and an instant preview of results (3).}
\label{fig:thegamma}
\end{figure}

@@

% TODO REWRITE

\item We show that iterative prompting is a complete and correct program construction method and
  consider how it lowers the barrier to entry and allows learning through examples, without
  expert guidance (Section~\ref{sec:design}).

@@

Our second type provider allows users to construct queries to explore data in tabular formats
such as CSV files. Unlike the data cube provider, the provider for tabular data does not just
allow selecting a subset of the data, but it can be used to construct SQL-like query. Consider
the example from Figure~\ref{fig:thegamma}:

@@

\begin{figure}
\centering
\begin{subfigure}[b]{0.5\textwidth}
  \centering
  \includegraphics[scale=0.25]{figures/cubetp}
  \vspace{0.5em}
  \caption{Exploring World Bank data using the data cube type provider, users
    choose values from two dimensions to obtain a data series.}
  \label{fig:cubetp}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.45\textwidth}
  \centering
  \includegraphics[scale=0.28]{figures/graphtp}
  \caption{To query graph data, the user specifies a path through the data, possibly with
    placeholders to select multiple nodes.}
  \label{fig:graphtp}
\end{subfigure}
\vspace{-0.5em}
\caption{Design of type providers for exploring data cubes and graph databases.}
\label{fig:three graphs}
\end{figure}

@@

\begin{thegamma}
drwho.Character.Doctor
  .'ENEMY OF'.'[any]'
  .'APPEARED IN'.'Day of the Moon'
\end{thegamma}

@@

\section{Design}
\label{sec:design}

\subsection{Journalism}

We start with a careful consideration of our target
application domain, i.e.~data analyses produced by journalists and citizen data scientists that
are published online. We look at both practical requirements for such programming environment
and requirements arising from our focus on journalism. This analysis is based on the author's
experience of collaborating with journalists\footnote{Citations removed to preserve anonymity.},
review of literature on data journalism, e.g.~\cite{ddj,edcj17,edcj18} and more general trends in
journalism.

% Open Journalism

Journalism continually develops and responds to the many challenges it faces \cite{future}.
Two recent challenges are relevant to our work. The first is building trust in media.
One way of establishing trust in the age of fake news is to be more transparent about editorial
decisions, process and original sources.

Many journalists believe that opening up the process
shows the quality and trustworthiness of their work~\cite{transparency}.
The second challenge is reader engagement. To develop a relationship with readers, journalists are
increasingly looking for meaningful ways of engagement. This includes reader comments, involvement
of citizen journalists \cite{comments,citizen} and the development of new interactive formats
\cite{youdraw}. To address the above challenges, a tool for data exploration should satisfy the
following three requirements.

\paragraph{Trust Through Transparency}
To support trustworthiness, data analyses should be transparent. The reader should be able to
determine what is the source of analysed data and how has the data been transformed. As much as
possible, these capabilities should also be accessible to non-expert readers.

\paragraph{Reproducibility for Fact Checking}
It should be possible to re-run the analysis to verify that it produces the presented results.
However, running an opaque script is not enough. A reader should be able to recreate the analysis
by following the necessary steps from the original data source to the end result.

\paragraph{Encouraging Meaningful Engagement}
The tool should support a mechanism through which readers can engage in a meaningful discussion.
For example, it should allow modifying of parameters of a data visualization in order to show
how different choices affect the final result.

\subsection{Lowering barriers}
Our aim is to make programmatic data exploration accessible to journalists, but we want to keep
the desirable properties of text-based programming. In particular, source code of a data
exploration should provide a full reproducible record of how the data analysis has been done.
As end-users, journalists have a number of interesting characteristics. They work under tight
deadline and data exploration is only a complementary skill. They also need to work with a wide
range of data sources, including big data tables (e.g.~Iraq War documents leak) or graph
databases (e.g.~Panama Papers). This leads to a number of practical requirements on the programming
environment.

\paragraph{Conceptual Simplicity}
We target end-users who cannot dedicate much time to learning about a tool prior
to using it. Consequently, using the tool should require understanding of only a small number
of concepts. Once the user understand a small number of concepts, they should be able to complete
basic data exploration tasks.

\paragraph{Uniformity across Data Sources}
The users should be able to navigate through large databases, query relational databases and
query graph databases through the same mechanism. Ideally, expertise gained with one data source
should also be transferable to working with another data source.

Data exploration tasks, such as querying tables, have a certain irreducible complexity.
Regardless of the interface, the user will be exposed to concepts such as filtering,
sorting and grouping. Our design aims to lower the barrier to entry by stratifying the concepts
into first-level \emph{iterative prompting} principle and second-level \emph{domain specific
languages} for each kind of data source.
The user needs to master the iterative prompting principle in order to start working
with the system. However, using individual data sources can be mastered gradually.

\paragraph{Iterative Prompting}
%Iterative prompting is an interaction principle where the user repeatedly invokes an
%auto-complete prompt and makes a selection from the offered options. The principle is closely
%related to both code completion and the use of command line. Iterative prompting is novel as an
%overarching interaction principle for program construction.

In code completion, the auto-complete
prompt is invoked only in certain contexts, e.g.~when accessing a member of an object through a
variable defined earlier in code. It requires the user to be sufficiently familiar with the
programming language in order to get to a point where they are offered a list of members. In
contrast, our system only requires choosing the initial data source. The rest of the programming
is done via iterative prompting. The iterative nature of iterative prompting makes it similar to
using the command line or REPL (read-eval-print-loop) tools. However, those repeatedly ask the
user to type code or commands.

%https://www.nngroup.com/articles/recognition-and-recall/
We argue that iterative prompting is easier than other forms of interaction, because it
follows the \emph{recognition over recall} usability heuristic. The users are only required to
choose from an offered list of options, rather than having to recall a possible command
(to type in a command line) or a syntax in a text-based programming environment.

\paragraph{Domain Specific Languages}
The access to individual data sources in The Gamma is facilitated through a domain specific
language, which defines the primitives that are offered to the user in the auto-complete prompts
invoked through iterative prompting. The domain specific languages are embedded in The Gamma --
they define merely the available members (and specify how to evaluate a chain of members), but
they cannot define any custom syntax.

We discuss the individual domain specific languages for querying data tables, graph databases and
data cubes in a later section. The complexity of those languages differs. The previously discussed
language for querying tabular data is the most complex one. However, The Gamma makes it easy for
the user to start exploring and learning new languages, because they are all accessible via
iterative prompting.

\subsection{Learnign without experts}

Sarkar \cite{learning} reports that users learn how to use Excel either by talking to experts,
or by seeing a feature in a spreadsheet received from a colleague. In our circumstances, experts
are unlikely to be available, so the tool should support learning from examples. When looking at
a work done and published by another person, the user needs to see (and be able to understand)
how a task was completed.

As discussed earlier, The Gamma is designed for users who work under tight deadlines and only
analyse data as their secondary task. The Gamma aims to support such users by having a low
barrier for entry and making it easy to learn independently.

\paragraph{Design for Percolation}
When analysing how Excel users learn Sarkar \cite{learning} points out that users learn new
features opportunistically when the usage of a feature is apparent in a spreadsheet. For
example, users can learn different functions to use in formulas, because those are visible in
the cell. Learning how to use a wizard for creating charts in this way is not possible because
it leaves no trace in the spreadsheet -- only the final result. Sarkar's recommendation is to
\emph{design for percolation}, i.e.~in a way where looking at the final result makes it apparent
what feature has been used and how.

\paragraph{Text-based Source Code}
In The Gamma, each step in the iterative prompting process results in an identifier that is
added to the source code. This means that a program constructed solely through iterative prompting
keeps a full trace of how it was created. Seeing the resulting source code provides the user all
information that they need to recreate the program, not just by copying it, but also by using the
iterative prompting mechanism. The Gamma represents code as text and allows the user to edit it
freely, so not all interactions leave a trace. For example, deleting the most recently selected
option and choosing a different one is not apparent from the result.

\subsection{Correct and complete}
any program can be created by autocomplete, all programs created by autocomplete are correct

@@

\newpage

% \begin{figure}
% \includegraphics[width=1\columnwidth]{figures/nyt}
% \caption{New York Times article on Obama's legacy \cite{youdraw}. The article asks the reader to make a guess
% (engagement), but only lists ``Immigration and Customs Enforcement'' as a source of data.}
% \label{fig:nyt}
% \end{figure}

% \begin{figure}
% \includegraphics[width=1\columnwidth]{figures/ft}
% \caption{Financial Times analysis of recyclable waste. Full source is provided as Jupyter Notebooks
% on GitHub \cite{ftnotebooks}, but re-running the analysis is difficult, even for an expert.}
% \label{fig:ft}
% \end{figure}

% ==================================================================================================
~
\newpage

@@

\begin{figure}
\includegraphics[width=1\columnwidth]{figures/sidebyside}
\caption{The Gamma environment with an automatically generated chart comparing CO2 emissions
of China and United States.}
\label{fig:sidebyside}
\end{figure}

\section{Use cases}
\label{sec:cases}

The Gamma aims to simplify programmatic data exploration while keeping enough expressive power
to allow users to create interesting data visualizations. We consider the expressive power in
this section and assess simplicity in the next one. We used The Gamma to analyse
the UK government expenditure, activities of a research institute, Olympic medals and
information about the Doctor Who series\footnote{Links removed to preserve anonymity.}.
In this section, we consider two interesting larger examples.

\subsection{If Michael Phelps were a Country}
Michael Phelps has won so many medals that media compared him to
countries.\footnote{\href{https://www.npr.org/sections/thetorch/2016/08/14/489832779/}{\small\bf\ttfamily npr.org/sections/thetorch/2016/08/14/489832779/}}
Using the tabular data provider, we construct a chart shown in Figure~\ref{fig:cases} that
produces a country league table including Michael Phelps:\footnote{Available at: \url{http://gallery.thegamma.net/86/}}

\begin{thegamma}
\kvd{let} data = olympics
    .'group data'.'by Team'.'sum Gold'.then
    .'sort data'.'by Gold descending'.then
    .paging.skip(43).take(4)
    .'get series'.'with key Team'.'and value Gold'

\kvd{let} phelps = olympics
  .'filter data'.'Athlete is'.'Michael Phelps'.then
  .'group data'.'by Athlete'.'sum Gold'.then
  .'get series'.'with key Athlete'.'and value Gold'

charts.bar(data.append(phelps))
  .setColors(["#aec7e8","#aec7e8","#1f77b4"])
\end{thegamma}

The core of the data analysis is done in two commands. The first counts gold medals by countries
and uses paging to fetch 4 countries with suitable number of medals. In the second, we abuse the
grouping operation to aggregate data for just a single group. The two data series are then assigned
to local variables (for readability) and passed to the \ikvd{chart.columns} function.
In this case study, the core data exploration can be constructed via iterative prompting, but
building a chart requires invoking the \ikvd{append} operation and specifying colors using a list.

\subsection{The Most Frequent Doctor Who Villains}
In the second case study, we list Doctor Who villains by the number of episodes in which they
appear.\footnote{Available at: \url{http://gallery.thegamma.net/87/}} This is interesting as it
combines the graph database provider for fetching the data with the tabular data provider for
summarization:

\begin{thegamma}
\kvd{let} topEnemies = drWho.Character.Doctor
  .'ENEMY OF'.'[any]'.'APPEARED IN'.'[any]'.explore
  .'group data'.'by Character name'
    .'count distinct Episode name'.then
  .'sort data'.'by Episode name descending'.then
  .paging.take(8).'get series'
    .'with key Character name'.'and value Episode name'

chart.bar(topEnemies)
\end{thegamma}

Lines 1-2 use the graph database provider to find all paths linking the Doctor node with any
character linked via \ikvd{ENEMY OF}, followed by any episode linked by \ikvd{APPEARED IN}.
This produces a table that can be analysed using the tabular data provider by choosing the
\ikvd{explore} member. For each character (the villain) we count the total number of
distinct episodes in the table. The result is shown in Figure~\ref{fig:cases}.

The logic is not trivial, but it performs a fairly sophisticated data analysis that involves a
graph database query followed by an SQL-like data aggregation. The code can be constructed
using iterative prompting (with the exception of the numbers in paging). It can also be constructed
gradually and instantaneous prviews support the user while doing so.

\begin{figure}
\centering
\includegraphics[width=1\columnwidth]{figures/cases}
\caption{Charts produced by two The Gamma case studies.}
\label{fig:cases}
\end{figure}

\section{User study}
\label{sec:study}

Our goal is to develop an easy-to-learn tool that journalists and other non-programmers can use
for producing transparent data analyses. To evaluate the extent to which The Gamma achieves this,
we conduct a user study. We recruit participants among the operations and business team of a
non-profit research organization and investigate three research questions.

\paragraph{RQ1: Can non-programmers explore data with The Gamma?}
We give participants one of four simple data exploration tasks and see if they can complete the
task using The Gamma and, possibly, how much assistance they need.

\paragraph{RQ2: Can knowledge transfer between data sources?}
Our first hypothesis is that users familiar with the iterative prompting principle will be able to
use an unfamiliar data source. In two of the tasks, participants are shown one data source and
then asked to work with a different one.

\paragraph{RQ3: Can users learn from just code samples?}
Our second hypothesis is that users can learn through percolation, i.e.~by looking at the source
code of published analyses. In one of our tasks, participants are given a code sample, but only a
minimal explanation of The Gamma environment.

\subsection{Study design}
We perform between-subjects study to evaluate the first experience of using The Gamma.
We recruit 13 participants (5 male, 8 female) from a business team
of a research lab working in various non-technical roles (project management,
partnerships, communications) including one former journalist.

We split participants into 4 groups. We first gave a brief 5 minute demonstration of The
Gamma and then asked participants to complete a task. Finally, we conducted a short semi-structured
group interview and later sent participants a follow-up questionnaire. The four tasks were:

\begin{itemize}
\item \emph{Expenditure.} Participants were given a demo using \emph{worldbank}.
  They were asked to use the \emph{expenditure} data source to compare the UK government spending
  on ``Public order and safety'' and ``Defence'' in terms of GDP.
\item \emph{Lodrs.} Participants were given a demo using \emph{worldbank}.
  They were asked to use the \emph{lords} data source (a table with House of Lords
  members expenses) to find a members representing London with the highest travel costs.
\item \emph{Worldbank.} Participants were given a minimal explanation of The Gamma environment and
  a code sample using \emph{worldbank}. They were asked to solve a different task using
  the \emph{worldbank} data source.
\item \emph{Olympics.} Participnts were given a demo using \emph{olympics}.
  They were asked to solve a more complex problem using the same data source.
\end{itemize}

@@

@@

\begin{table}
  \centering
  \begin{tabular}{p{19em} c c}
    \toprule
      {\small \textit{Questionnaire item}} & {\small \textit{Avg}} & {\small \textit{Sdv}} \\
    \midrule
    \small Considering your experience \& possible use of The Gamma:\\
    \small \quad I found the system easy to use. & \small 2.11 & \small 0.87\\
    \small \quad Journalists will be able to use it to analyse data. & \small 2.67 & \small 1.05\\
    \small \quad Readers will be able to critically examine the analyses. & \small 2.44 & \small 0.68\\
    \small Thinking about the usability of the system:\\
    \small \quad The resulting programs are easy to understand. & \small 2.33 & \small 0.82\\
    \small \quad The text editor for creating them was easy to use. & \small 2.33 & \small 0.94\\
    \small \quad Finding the right item in the auto-complete was easy. & \small 2.56 & \small 1.42\\
    \small I would know how to approach a task using a new data source\\
    \small \quad If given a comprehensive video tutorial. & \small 2.00 & \small 1.24\\
    \small \quad If given a number of code samples. & \small 2.78 & \small 1.13\\
    \small \quad Without any further guidance. & \small 3.56 & \small 1.17\\
    \bottomrule
  \end{tabular}
  \caption{Summary of follow-up questionnaire responses using
    a 5-point Likert scale (1 = strongly agree, 5 = strongly disagree) over 9 subjects.}
  \label{tab:quest}
\end{table}

@@

\section{Conclusions}
We present The Gamma, a simple data exploration environment. The Gamma is motivated by the needs
of journalists working with data, who need to produce transparent and easy-to-understand reports
backed by data that are transparent and engaging, all the while facing tight deadlines.

>>>>

% Explain type providers a bit more (we say what it does, but not much about what
% it is - give dynamic typing in a static language)
%
% The 'then' operation was mysterious - can we add one line explanation to the operations?
% (yes we can and we should totally do that)

% \usepackage{xspace}
\usepackage{fancyvrb}
% \usepackage{inconsolata}
\usepackage{xcolor}
\usepackage{tikz}
\newcommand*{\priority}[1]{\begin{tikzpicture}[scale=0.12]%
   \draw (0,0) circle (1);
   \fill[fill opacity=1,fill=black] (0,0) -- (90:1) arc (90:90-#1*3.6:1) -- cycle;
   \end{tikzpicture}}
\newcommand*{\priorityc}[1]{
\resizebox{0.8em}{0.8em}{
 {\protect\tikz{ \protect
   \draw[line width=1mm, black] (0,0) circle (1); \fill[fill opacity=1,fill=black] (0,0) -- (90:1) arc (90:90-#1*3.6:1) -- cycle;
   } }}}
%
\definecolor{kvdclr}{rgb}{0.3,0.0,0.8}
\DefineVerbatimEnvironment{thegamma}{Verbatim}{fontfamily=zi4,numbers=left,xleftmargin=6mm,fontsize=\small,commandchars=\\\{\}}
% %\newcommand{\kvd}[1]{\textcolor{kvdclr}{#1}}
\newcommand{\kvd}[1]{\textbf{#1}}
\newcommand{\ikvd}[1]{{\fontfamily{zi4}\selectfont\small #1}}
% \usepackage{balance}       % to better equalize the last page
% \usepackage{graphics}      % for EPS, load graphicx instead
% \usepackage[T1]{fontenc}   % for umlauts and other diaeresis
% \usepackage{txfonts}
% \usepackage{mathptmx}
% \usepackage[pdflang={en-US},pdftex]{hyperref}
% \usepackage{color}
% \usepackage{booktabs}
% \usepackage{textcomp}

@@

\begin{figure}
\includegraphics[width=.58\columnwidth]{figures/thegamma-annot}
\vspace{-0.5em}
\caption{Obtaining teams with the greatest number of gold medals from Rio 2016
Olympics with a reproducible The Gamma script (1), contextual iterative prompting mechainsm
offering ways of sorting the data (2) and an instant preview of results (3).}
\label{fig:thegamma}
\vspace{-0.5em}
\end{figure}

@@

\item We reflect how our design addresses challenges faced by journalists (Section~\ref{sec:design-ddj})
  and analyse the design theoretically. We argue that our design lowers barriers to entry (Section~\ref{sec:design-bar}),
  supports learning without experts (Section~\ref{sec:design-expert}) and offers a
  complete and correct program construction method (Section~\ref{sec:design-cc}).

@@

Our second type provider allows users to construct queries to explore data in tabular formats
such as CSV files. A prominent example of tabular data used by journalists is, e.g.~the
Iraq War documents leak~\cite{iraq}. Unlike the data cube provider, the provider for tabular data does not just
allow selecting a subset of the data, but it can be used to construct SQL-like query. Consider
the example from Figure~\ref{fig:thegamma}:

@@

\begin{figure}
\centering
\begin{subfigure}[b]{0.5\textwidth}
  \centering
  \includegraphics[scale=0.25]{figures/cubetp}
  \vspace{0.5em}
  \caption{Exploring World Bank data using the data cube type provider, users
    choose values from two dimensions to obtain a data series.}
  \label{fig:cubetp}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.45\textwidth}
  \centering
  \includegraphics[scale=0.28]{figures/graphtp}
  \caption{To query graph data, the user specifies a path through the data, possibly with
    placeholders to select multiple nodes.}
  \label{fig:graphtp}
\end{subfigure}
\vspace{-0.5em}
\caption{Design of type providers for exploring data cubes and graph databases.}
\label{fig:tps}
\end{figure}

@@

\begin{thegamma}
drwho.Character.Doctor
  .'ENEMY OF'.'[any]'.'APPEARED IN'.'Day of the Moon'
\end{thegamma}

@@

\section{Reflections on Design}
\label{sec:design}

The design of The Gamma has been motivated by a curiosity as to whether iterative
prompting can make text-based programming with data accessible to non-experts. In this section,
we theoretically assess the resulting design and position it in the context of a possible application
in the context of data journalism.

\subsection{Data Journalism Perspective}
\label{sec:design-ddj}

Although our work targets non-expert data exploration in general, data journalism is an important
specific domain with interesting design requirements. We reflect those on the basis of relevant
literature, e.g.~\cite{ddj,edcj17,edcj18} and past experience of collaborating with
journalists. Two challenges faced by journalists \cite{future} are particularly relevant to The Gamma.

\subsubsection*{Trust Through Transparency}
The first issue is trust in media. Many practitioners believe that transparency
about the editorial process and information sources can serve as a proof of quality and
trustworthiness~\cite{transparency}. This applies to data analyses too. For example, the Financial Times
increasingly share source code of their analyses on GitHub, e.g.~\cite{ftnotebooks},
but reproducing such analyses is difficult even for an expert. The Gamma has a potential to
improve on the status quo in that a part of the analysis can be directly embedded in an article.
The source code provides full account of what the data sources are and how are they used.
The Gamma makes this code, to a greater extent, accessible to non-expert readers.

\subsubsection*{Encouraging Meaningful Engagement}

The second challenge is developing relationship with readers. Journalists are looking
for meaningful ways of engagement through reader comments, involvement of citizen journalists
\cite{comments,citizen} and the development of new interactive formats \cite{youdraw}.
The Gamma can support this aim by providing a data exploration environment that can engage
non-expert readers in a meaningful discussion. As argued below, non-experts can learn to use the
system and can, for example, modify parameters or change filtering criteria to get a
new perspective on a story.

\subsection{Lowering Barriers to Entry}
\label{sec:design-bar}

Data exploration has a certain irreducible essential complexity. To make it accessible to users who
cannot dedicate much time to learning a tool prior to using it, this complexity needs to be
carefully stratified. The Gamma uses a two-level structure. At the first level, the user needs
to learn iterative prompting to be able to start exploring data. At the second level, users will
need to learn domain-specific languages defined by individual type providers.

\subsubsection*{Iterative Prompting Principle}
Iterative prompting is a suitable first level principle, because it is easy to trigger.
In conventional programming languages, auto-complete assistance is only available once basic
code structure is written. In contrast, user of In The Gamma only needs to choose the initial
data source. Iterative prompting is also easier to use than e.g.~a command line or a REPL
(read-eval-print-loop), because it follows the \emph{recognition over recall} usability heuristic.
The users are not required to recall and type a command. They merely need to select one from a
list of options.

\subsubsection*{Stratifying Data Exploration Complexity}
At the first level, any data source is accessed through the same mechanism. At the second level,
each data source defines its own domain specific language, consisting of the primitives that are
offered to the user in auto-complete prompts. The domain specific languages are embedded in The Gamma
language -- they define merely the available members and cannot extend the core language syntax.
Although the complexity of individual languages differs, the users can always start exploring
and learning new languages using the familiar first level iterative prompting principle.
An important question, which we study in Section~\ref{sec:study}, is whether the expertise gained
with one data source be transferred to working with another data source.

\subsection{Learning without Experts}
\label{sec:design-expert}

A typical user of The Gamma will not dedicate significant time to learning it in advance and
they will not have access to experts. Most learning thus needs to happen from examples.
When analysing how Excel users learn, Sarkar \cite{learning} points out that users learn new
features when the usage of a feature is apparent in a spreadsheet. For example, users can learn
different functions in formulas, because those are visible in the cell. Learning how to use a
wizard for creating charts is not possible because the operation leaves no full trace in the
spreadsheet. Sarkar's recommendation is to \emph{design for percolation},
i.e.~in a way where looking at the final result makes it apparent what feature has been used and how.

In The Gamma, each step of iterative prompting results in an identifier that is
added to the source code. This means that a program constructed solely through iterative prompting
keeps a full trace of how it was created. The resulting source code provides the user all
information that they need to recreate the program, not just by copying it, but also by using the
iterative prompting mechanism. We assess the viability of this way of learning in the study in
Section~\ref{sec:study}.

\subsection{Correctness and Completeness}
\label{sec:design-cc}

An important characteristic of our design is that, barring a few exceptions discussed below, the iterative
prompting mechanism is both correct and complete with respect to possible data exploration scripts.
This means that (i) auto-complete lists offered by iterative prompting contain
only options that are valid in a given context and that (ii) any script that can be written
in The Gamma can be contructed via iterative prompting.

\subsubsection*{Correctness} The Gamma uses an object-based type system for error-checking and for
generating auto-complete lists. A well-typed script can always be executed. When using iterative
prompting, a selected option, which is a valid object member, is added to the end of a script,
leading to a well-typed script. This distinguishes our system from auto-completion based on machine
learning or dictionary-based methods, which may offer members not valid in a given context. Advanced
users of The Gamma, modifying scripts as text, can still violate correctness -- for example, if the
user uses auto-complete to replace a member in a middle of a member access chain with a member of another type.

\subsubsection*{Completeness} Auto-completion lists offered via iterative prompting contain
all available members and so the user can construct all possible scripts. The range of such scripts
is determined by type providers, which support limited set of options (e.g. group aggregations in
the tabular data type provider). Two exceptions to completeness in our current design is specifying
numerical parameters as in \ikvd{take(5)} and advanced language features such as let binding.
We believe that alleviating the need for those poses an interesting further research challenge.

\begin{figure}[b]
\centering
\begin{subfigure}[b]{0.49\textwidth}
  \centering
  \includegraphics[width=1\columnwidth]{figures/cases-mp}
  \caption{Exploring Olympic medalists using tabular data type provider}
  \label{fig:cases-mp}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.49\textwidth}
  \centering
  \includegraphics[width=1\columnwidth]{figures/cases-dr}
  \caption{Exploring Dr Who graph database by composing type providers}
  \label{fig:cases-dr}
\end{subfigure}
\vspace{-0.5em}
\caption{Charts produced by two case studies of using The Gamma.}
\label{fig:cases}
\end{figure}

@@

\section{Case Studies}
\label{sec:cases}

The Gamma aims to simplify programmatic data exploration while keeping enough expressive power
to allow users to create interesting data visualizations. We consider the expressive power in
this section and assess simplicity in the next one. We used The Gamma to analyse
the UK government expenditure, activities of a research institute, Olympic medals and
information about the Doctor Who series\footnote{The analyses are available (non-anonymously)
at \url{http://gallery.thegamma.net}, \url{http://turing.thegamma.net} and \url{http://rio2016.thegamma.net}}.
In this section, we consider two interesting larger examples\footnote{Available
(non-anonymously) at: \url{http://gallery.thegamma.net/86/} and  \url{http://gallery.thegamma.net/87/}, respectively.}.

\subsubsection*{If Michael Phelps were a Country}
Michael Phelps has won so many medals that media compared the number to
countries.\footnote{e.g.~``If Michael Phelps Were A Country, Where Would His Gold Medal Tally Rank?''
 \url{https://www.npr.org/sections/thetorch/2016/08/14/489832779/}}
Media illustrated the story with a chart showing a country league table including Michael Phelps.
We reproduce the chart, shown in Figure~\ref{fig:cases-mp}, using the tabular data type provider:

\begin{thegamma}
\kvd{let} data = olympics.'group data'.'by Team'.'sum Gold'.then
    .'sort data'.'by Gold descending'.then
    .paging.skip(43).take(4).'get series'.'with key Team'.'and value Gold'

\kvd{let} phelps = olympics.'filter data'.'Athlete is'.'Michael Phelps'.then
  .'group data'.'by Athlete'.'sum Gold'.then
  .'get series'.'with key Athlete'.'and value Gold'

charts.bar(data.append(phelps)).setColors(["#aec7e8","#aec7e8","#1f77b4"])
\end{thegamma}

\noindent
The data analysis is done in two commands. The first counts gold medals by countries
and uses paging to fetch 4 countries with suitable number of medals. In the second, we abuse the
grouping operation to aggregate data for just a single group. The two data series are then assigned
to local variables (for readability) and passed to the \ikvd{chart.columns} function.
The case study illustrates when more advanced language features are necessary. The data exploration
itself has been completed via iterative prompting, but producing the final chart currently requires
some manual programming. In practice, this would likely be done with the help of an expert
or by copying code from an example.

\subsubsection*{The Most Frequent Doctor Who Villains}
In the second case study, use a graph database on the Dr Who series and list Dr Who villains
by the number of episodes in which they appear. This case study is interesting as it
combines the graph database provider for fetching the data with the tabular data provider for
summarization:

\begin{thegamma}
drWho.Character.Doctor.'ENEMY OF'.'[any]'.'APPEARED IN'.'[any]'.explore
  .'group data'.'by Character name'.'count distinct Episode name'.then
  .'sort data'.'by Episode name descending'.then
  .paging.take(8).'get series'.'with key Character name'.'and value Episode name'
\end{thegamma}

\noindent
Line 1 use the graph database provider to find all paths linking the Doctor node with any
character linked via \ikvd{ENEMY OF}, followed by any episode linked by \ikvd{APPEARED IN}.
This produces a table that can be analysed using the tabular data provider by choosing the
\ikvd{explore} member. For each character (the villain) we count the total number of
distinct episodes in the table. The result is shown in Figure~\ref{fig:cases-dr}.
The example performs a fairly sophisticated data analysis that involves a graph database query
followed by an SQL-like data aggregation. The code can be fully constructed using iterative
prompting (with the exception of the numbers in paging). Another important feature is that
instantaneous previews support the user during the construction, showing the intermediate result
at each step of the construction.

@@

\section{User Study}
\label{sec:study}

Our goal was to develop an easy-to-learn tool that non-programmers can use
for producing transparent data analyses. To evaluate the extent to which The Gamma achieves this,
we conduct a user study. We recruit participants among the operations and business team of a
non-profit research organization and investigate three research questions.

\vspace{0.5em}\noindent\emph{RQ1: Can non-programmers explore data with The Gamma?}
We give participants one of four simple data exploration tasks and see if they can complete the
task using The Gamma and, possibly, how much assistance they need.

\vspace{0.5em}\noindent\emph{RQ2: Can knowledge transfer between data sources?}
Our first hypothesis is that users familiar with the iterative prompting principle will be able to
use an unfamiliar data source. In two of the tasks, participants are shown one data source and
then asked to work with a different one.

\vspace{0.5em}\noindent\emph{RQ3: Can users learn from just code samples?}
Our second hypothesis is that users can learn through percolation, i.e.~by looking at the source
code of published analyses. In one of our tasks, participants are given a code sample, but only a
minimal explanation of The Gamma environment.

\subsection{Study design}
We perform between-subjects study to evaluate the first experience of using The Gamma.
We recruit 13 participants (5 male, 8 female) from a business team
of a research lab working in non-technical roles (project management,
partnerships, communications) including one former journalist. Only one participant (\#12)
had prior programming experience.

We split participants into 4 groups. We first gave a brief 5 minute overview of The
Gamma and then asked participants to complete a task.
 The four tasks were:

\begin{itemize}
\item \emph{Expenditure.} Participants were given a demo using \emph{worldbank}.
  They were asked to use the \emph{expenditure} data source to compare the UK government spending
  on ``Public order and safety'' and ``Defence'' in terms of GDP.
\item \emph{Lords.} Participants were given a demo using \emph{worldbank}.
  They were asked to use the \emph{lords} data source (a table with House of Lords
  members expenses) to find a members representing London with the highest travel costs.
\item \emph{Worldbank.} Participants were given a minimal explanation of The Gamma environment and
  a code sample using \emph{worldbank}. They were asked to solve a different task using
  the \emph{worldbank} data source.
\item \emph{Olympics.} Participnts were given a demo using \emph{olympics}.
  They were asked to solve a more complex problem using the same data source.
\end{itemize}

Finally, we conducted a short semi-structured
group interview and later sent participants a follow-up questionnaire.

@@

x

y

z

a

b

c

d

@@

@@

\section{Conclusions}
[TODO: THIS IS A SHORT PAPER.... no design probe]

We present The Gamma, a simple data exploration environment. The Gamma is motivated by the needs
of journalists working with data, who need to produce transparent and easy-to-understand reports
backed by data that are transparent and engaging, all the while facing tight deadlines.
